A generalization of the Bradley-Terry model for draws in chess

### 2019-05-20 inst/collusion_chess_draws.Rnw
further tweaking

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess game is
regarded as a three-way competition between the two players and an
entity that wins if the game is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude, on the basis
of a dataset that includes only the scorelines at the event itself,
that the scoreline of the Candidates Tournament (Cura\c{c}ao 1962)
exhibits evidence of collusion ($p<10^{-5}$), in agreement with
previous work that uses additional information such as historical
results including number of moves.}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
idenfied as the winner, but the system has been applied to many
non-competitive situations such mating preference~\citep{west2008} and
consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
considers $n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way choices such as mating preferences~\citep{hankin2010}.

@@

Such arguments suggest considering a drawn chess game as a
qualitatively different outcome from a statement of equality of skill:
the play itself tends to be different, the drawing players have a
different mindset, and drawn games have less value in the wider
context of the game.

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as
a three-way competition between the two players and an entity that
wins if the game is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.

If we consider a chess game to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,D$ with
$D+\sum p_i=1$ and this would be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity, the White Monster.
Recalling that a game might be a draw, we obtain the probabilities as
given in Table~\ref{probabilities}.  The corresponding likelihood
function is given below in equation \ref{likelihood_WD}:

@@

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets.  The general
approach follows~\citet{edwards1972}.  To test any hypothesis $H$ on a
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

This likelihood function is easily maximized numerically
and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing white: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

The method of support allows one to estimate proclivity to draw while
accounting for players' strengths and first-move advantage.  We now
consider one famous incident, following the World Chess Championships
1963, in which Fischer famously and publicly claimed that the Soviet
players colluded.  The accusation was that Soviet players agreed ahead
of time to draw when playing one another, in order to conserve energy
for subsequent matches.

@@

Here I analyse the original subject of the complaint, the Cura\c{c}ao
1962 Candidates (match results shown in Table~\ref{candidates}).
Fischer alleged that three of the Soviet players (Keres, Petrosian,
and Geller) had agreed to draw amongst themselves.  Observe that all
twelve of the games between the three accused players ended in a draw,
a significant result given 51 winning games and 54 draws (Fisher exact
test, one-sided, $p=0.0079$).

To investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet draw monster $C$, who wins when two colluding
players draw, and is inactive whenever a game has at least one
non-colluding player.  Then $H_N\colon C=D$ corresponds to no
collusion and the likelihood function would be

@@

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] on the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

The fith interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
table~\ref{interzonal}, we would have a likelihood function

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster $D_s$, who wins when two Soviet
players draw, and is inactive if a game has at least one non-Soviet
player.  Then $H_N\colon D_s=D$ corresponds to no collusion and the
likelihood function for table~\ref{interzonal} would be

>>>>

In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing an entity whose strength is added to that
of the white player.  These techniques afford insight into players'
strengths, response to playing black or white, and risk-aversion as
manifested by probability of drawing.  The likelihood functions
arising are easily optimised numerically.  I analyse a number of
datasets of chess results, including the infamous World Chess
Championships 1963 in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way competitions such as {\em MasterChef
  Australia}~\citep{hankin2017}, and the system has been applied to
many non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

@@

Such arguments suggest that one should infer more from a drawn game
than simple equality of skill: the play itself tends to be different
and more cautious, the drawing players have a different mindset, and
drawn games have less value in the wider context of the game.

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and an entity that wins
if the game is drawn.  The system is further modified to account for
first-move advantage by positing an entity whose strength is added to
that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  If we consider
a chess game to be a three-way trial, an appropriate likelihood
function for non-negative $p_1,\ldots,p_n,D$ with $D+\sum p_i=1$ would
be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity with strength $W$.  Recalling
that a game might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets using the Method
of Support~\citet{edwards1972}.  To test any hypothesis $H$ on
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

This likelihood function is easily maximized numerically and this is
shown visually in figure~\ref{maxp_karpov_kasparov_anand}, which
suggests a number of interesting and testable statistical hypotheses.
Firstly, we observe that $D=0$ cannot possibly be the case: the
likelihood for this hypothesis would be identically zero, owing to the
nonzero term in the numerator of equation~\ref{kka_likelihood}.  From
the figure, we see that the maximum likelihood estimate for draw
strength $\hat{D}\simeq 0.53$: compare the na\"ive estimate of
$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
games.  The MLE is lower because the maximization process
automatically adjusts for the different players' strengths.  Secondly,
we may make inferences about $W$, the extra strength conferred by
playing white: it is conceivable that $W=0$ as this term does not
appear on its own in the numerator of any term.  We may test
$H_0\colon W=0$ using the Method of Support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

The Method of Support allows one to estimate proclivity to draw while
accounting for players' strengths and first-move advantage.  We now
consider one famous incident, following the World Chess Championships
1963, in which Fischer famously and publicly claimed that three of the
Soviet players colluded.  The accusation was that Keres, Petrosian,
and Geller agreed ahead of time to draw when playing one another, in
order to conserve energy for subsequent matches.

@@

Here I consider only the game results, reproduced in
table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

@@

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] on the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

The fifth Interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
table~\ref{interzonal}, we would have a likelihood function

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster with strength~$D_s$, who wins when two
Soviet players draw, and is inactive if a game has at least one
non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
and the likelihood function for table~\ref{interzonal} would be
